---
title: "Pipeline gallery"
format:
  html:
    toc: true
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
---

```{r}
#| label: setup
#| message: false
library(gdalviz)
```

Real-world `gdal vector pipeline` examples rendered with `render_reactflow()`.
Every diagram is live: drag to pan, scroll to zoom, use the minimap, and
**click any node** to open the inspector with that step's full arguments,
propagated stream state, and GDAL docs link.

## TIGER states: a linear ETL pipeline

A classic download-filter-reshape-write flow, straight from the US Census
TIGER shapefiles over `/vsizip/vsicurl/`. Note the edge badges tracking what
changes along the stream: the SQL step reduces the schema to 6 fields, and
`reproject` stamps the CRS.

```{r}
#| label: tiger
tiger <- paste(
  "gdal vector pipeline",
  '! read --input "/vsizip/vsicurl/https://www2.census.gov/geo/tiger/TIGER2025/STATE/tl_2025_us_state.zip/tl_2025_us_state.shp"',
  '--input-layer tl_2025_us_state --open-option ENCODING=UTF-8 --open-option ADJUST_GEOM_TYPE=FIRST_SHAPE',
  "! filter --where \"STATEFP NOT IN ('02','15','60','66','69','72','74','78')\"",
  '! sql --sql "SELECT GEOID AS geoid, STATEFP AS state_fips, STUSPS AS state_abbr,',
  'NAME AS state_name, ALAND AS area_land_m2, AWATER AS area_water_m2 FROM tl_2025_us_state"',
  "! make-valid",
  "! set-geom-type --multi --skip",
  "! sort --method hilbert",
  "! write --output tiger_states.parquet --output-format Parquet",
  '--layer-creation-option COMPRESSION=ZSTD --layer-creation-option GEOMETRY_ENCODING=GEOARROW'
)

render_reactflow(tiger, theme = "dark", height = 620)
```

## Tee branching: valid / invalid geometry triage

`tee` fans the feature stream out into side pipelines -- here splitting
parcels by `ST_IsValid()` into separate outputs while the main stream
continues to a Parquet write. Branch flows render dashed.

```{r}
#| label: tee
triage <- paste(
  "gdal vector pipeline",
  '! read --input parcels.gpkg --layer parcels_statefp13',
  '! tee',
  '--tee-pipeline [ sql --sql "SELECT * FROM parcels_statefp13 WHERE ST_IsValid(geom)" --dialect SQLITE',
  '! write --output valid_parcels.gpkg --overwrite ]',
  '--tee-pipeline [ sql --sql "SELECT * FROM parcels_statefp13 WHERE NOT ST_IsValid(geom)" --dialect SQLITE',
  '! check-geometry ! write --output invalid_parcels.gpkg --overwrite ]',
  '! write --output all_parcels.parquet --output-format Parquet',
  '--layer-creation-option COMPRESSION=ZSTD'
)

render_reactflow(triage, direction = "LR", theme = "light", height = 540)
```

## Config-heavy runs and merged schema chains

A production parcels ETL: eight `--config` options group into the *runtime
configuration* node, thirty-two one-field-at-a-time `set-field-type` steps
(a GDAL nuance -- the command accepts one field per step) merge into a single
stacked `set-field-type` ×32 node, and because the pipeline writes GDALG, the
graph attaches the implicit *streamed output* sink. Click the merged node to
see every field -> type coercion.

```{r}
#| label: config-merged
fields <- c(
  source_fid = "Integer64", geoid = "String", state_fips = "String",
  county_fips = "String", tax_year = "Integer", num_buildings = "Integer",
  num_units = "Integer", year_built = "Integer", num_floors = "Integer",
  building_sqft = "Integer64", num_bedrooms = "Integer",
  num_half_baths = "Integer", num_full_baths = "Integer",
  improvement_value = "Integer64", land_value = "Integer64",
  ag_value = "Integer64", total_value = "Integer64", tax_acres = "Real",
  calc_area = "Real", sale_amount = "Real", sale_date = "Date",
  firm_date = "Date", static_bfe = "Real", elevation_min = "Real",
  elevation_max = "Real", elevation_avg = "Real", num_fireplaces = "Integer",
  updated = "Date", centroid_x = "Real", centroid_y = "Real",
  surface_point_x = "Real", surface_point_y = "Real"
)

parcels <- paste(
  "gdal vector pipeline",
  "--progress",
  "--config CPL_DEBUG=ON --config CPL_TIMESTAMP=ON --config CPL_LOG_ERRORS=ON",
  "--config CPL_LOG=logs/state_fips=44.gdal.log --config GDAL_NUM_THREADS=ALL_CPUS",
  "--config OGR_GPKG_NUM_THREADS=4 --config OGR_SQLITE_CACHE=4000000",
  "--config OGR_SQLITE_SYNCHRONOUS=OFF",
  "read --input parcels/LR_PARCEL_NATIONWIDE_FILE_US_2026_Q2.gpkg",
  "--input-layer lr_parcel_us --open-option LIST_ALL_TABLES=NO",
  '! filter --bbox "-71.907258,41.095834,-71.088571,42.018799" --where "statefp=\'44\'"',
  "! sql --sql @sql/gpkg.schema.full.sql --dialect SQLITE",
  "! make-valid",
  "! set-geom-type --multi --skip",
  paste(
    sprintf("! set-field-type --field-name %s --field-type %s", names(fields), fields),
    collapse = " "
  ),
  "! reproject --output-crs EPSG:4326"
)

render_reactflow(parcels, theme = "dark", height = 720)
```

## Pasted scripts just work

The same parser accepts pasted PowerShell (or bash) scripts with line
continuations -- handy when your pipelines live in `.ps1` files:

```{r}
#| label: script
ps1 <- 'gdal vector pipeline `
  read --input "G:/data/states.gpkg" --input-layer states `
! filter --where "STUSPS IN (\'GA\',\'FL\',\'AL\')" `
! reproject --output-crs EPSG:5070 `
! write --output southeast.fgb --output-format FlatGeobuf --overwrite'

render_reactflow(ps1, direction = "LR", theme = "light", height = 320)
```

## From diagram back to data

Every diagram above is backed by the same parsed model, so nothing here is
render-only. The graph is plain data:

```{r}
#| label: as-data
g <- pipeline_graph(triage)
g$nodes[, c("command", "category", "branch_role", "description")]
g$edges
```

and the pipeline serializes back to a canonical command line, a formatted
shell script, or a GDALG specification via `render_command_line()`,
`render_script()`, and `as_gdalg()`.
